Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-12-16 1.Paper/Paper.Rmd
update manuscript

Modelling data under MNAR mechanism implies to specify information about the missingness process in addition to assumptions about the observed data. A popular approach to address MNAR mechanism is selection models such as the one proposed by Heckman (1976). [@heckman1976] Briefly, the Heckman selection model corrects for selection bias by estimating two linked equations: a outcome equation, where the missing variable is associated with predictors, and a selection equation, that accounts for the inclusion of observations in the sample.   An important feature of the Heckman selection model is that it does not assume data to be MNAR, so that it can also be used when data are MCAR or MAR. It therefore offers an appealing solution to incomplete data sets when the missingness mechanism is not precisely known.

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life data generating mechanism may deviate a BVN distribution. <!-- This facilitates the Heckman model estimation and makes it more robust to deviations of distributional assumptions. -->

@@

 The Heckman model can be extended to hierarchical settings, i.e., in cases when individuals or sampling units are nested within clusters, as is the case in EHR or IPD. <!--In this case, sample units from the same cluster are expected to share some characteristics (e.g., the distribution of variables and correlations between them, relationships between exposure and outcomes, missing processes).

Since most statistical analyses assume that the sample units are independent of each other, more complex hierarchical models dealing with nested data are required. --> This hierarchical complexity must not only be taken into account in the analysis model, but also when dealing with missing data, thus requiring imputation models that are congenial to the analysis model, i.e. that make the same assumptions about the data.

Different procedures can be adopted to combine information between clusters; however, in our imputation method we opted for the two-stage approach that is often used in meta-analyses.[@simmonds2005] This is because such an approach is computationally less intensive and could potentially generate fewer convergence problems in the estimation of the Heckman hierarchical model compared to other approaches.

Briefly, in a first stage  we estimate $\boldsymbol{\theta_i}=\{\boldsymbol{\beta_i^O},\boldsymbol{\beta_i^S},\sigma_i,\rho_i\}$, for all $i$ between 1 and $N$ clusters, the cluster specific parameters of the Heckman model. That is, we estimate separately for each of the clusters, the parameters of the two equations in each study, the outcome model (\ref{eq:1}) and the selection model (\ref{eq:2}). In the second stage, we combine all $\boldsymbol{\theta_i}$  by using a random effects multivariate meta-analysis model.

In a random effects model, $\boldsymbol{\theta_i}$ parameters are assumed to be drawn independently and identically from a latent multivariate normal distribution of parameters with a  population mean $\boldsymbol{\Theta}$ and a population variance $\boldsymbol{\psi}$. [@higgins2009] Thus, we can write $\boldsymbol{\theta_i}=\boldsymbol{\Theta} + \boldsymbol{b_i}$  using $\boldsymbol{b_i} \sim N(0,\boldsymbol{\psi})$ random effects to allow for between-study heterogeneity in observed data relationships, and between-study heterogeneity in missing patterns.

# Using the Heckman model to impute missing data
\label{sec:methods}
We follow a similar approach proposed by Resche-Rigon and White (2018) [@resche-rigon2018] for multilevel imputation of data. Briefly, their method was developed to impute variables from a hierarchical structure (i.e., when there are units grouped within a cluster).
This method involves estimating an outcome equation (describing the relationships of the observed data to the missing variable) separately in each cluster, after which the parameter estimates of that equation are pooled using random effects meta-analysis.

With this method, values can be imputed in very common scenarios in IPD, e.g., sporadic and systematic missing patterns. In particular, when the response variable is systematically missing within a group, i.e., when $y_{ij}$ are totally missing within a cluster, the imputation values are drawn from a (generalized) linear model conditional on $\boldsymbol{\Theta}$ the marginal population parameters, i.e., those estimated after pooling the cluster-specific parameters $\boldsymbol{\theta_i}$. On the other hand, when the variable is sporadically missing within the cluster, i.e., there are some observed $y_{ij}$ within the cluster, the imputation model is conditional on the shrunk-cluster parameters, i.e., those coming from the shrinkage of $\boldsymbol{\theta_i}$ towards $\boldsymbol{\Theta}$.

Our imputation approach differs crucially from the previous approach as here we estimate two correlated equations (instead of a single outcome equation) in each cluster, thus obtaining $\boldsymbol{\theta_i}=\{\boldsymbol{\beta_i^O},\boldsymbol{\beta_i^S},\sigma_i,\rho_i\}$ parameters from both equations which are then pooled into a $\boldsymbol{\Theta}$ parameter set at the marginal or population level. Our imputation method was created for imputing datasets with a single missing outcome or covariate variable, but it can also be used to impute multiple incomplete variables in a data set.

@@

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\boldsymbol{\widehat{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. 
The Heckman model is estimated with the \textbf{gjrm} function of the GJRM R package[@radiceGJRMGeneralisedJoint2021],  under the bivariate model with the nonrandom sample selection (BSS) specification, from which we obtain not only the parameters' point estimates $\boldsymbol{\widehat{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

@@

In this step, we pool the parameters $\boldsymbol{\widehat{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation.

We also perform a univariate random effects meta_analysis on $\sigma'$, the log-transformed parameter of $\sigma$, and another on $\rho'$, the fisher-transformed parameter of $\rho$.

The meta-analysis model is performed with the \textbf{mixmeta} function of the R package mixmeta[@gasparrini2021], which allows the use of maximum likelihood (ML), restricted maximum likelihood (REML), and moments estimation methods. For the simulation and illustrative study, we use the restricted REML estimation method, which is recommended as it has a good balance between unbiasedness and efficiency. [@viechtbauer2005]

@@

From the meta-analysis model, we obtain the marginal estimates $\boldsymbol{\widehat{\Theta}}$ and the between-cluster variance matrix $\boldsymbol{\widehat{\psi}}$,i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\Theta}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\Theta^*}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

@@

@@

- \textbf{Bias:} Bias on the coefficient and random effect parameters.
- \textbf{Coverage:} Coverage of the 95\% confidence intervals for the coefficient  parameters.
- \textbf{Width}: Average width of the confidence interval on the coefficient parameters.
- \textbf{RMSE:} Root mean squared error of the coefficient  and random effect parameters.

>>>>

Modelling data under MNAR mechanism implies to specify information about the missingness process in addition to assumptions about the observed data. Two major approach have been used to address MNAR mechanism: pattern mixture models[@little1996] and selection models [@vella1998]. One of the most popular techniques within selection models is the one proposed byHeckman (1976). [@heckman1976] Briefly, the Heckman selection model corrects for selection bias by estimating two linked equations: a outcome equation, where the missing variable is associated with predictors, and a selection equation, that accounts for the inclusion of observations in the sample. An important feature of the Heckman selection model is that it does not assume data to be MNAR, so that it can also be used when data are MCAR or MAR. It therefore offers an appealing solution to incomplete data sets when the missingness mechanism is not precisely known.

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life data generating mechanism may deviate a BVN distribution.

@@

 The Heckman model can be extended to hierarchical settings, i.e., in cases when individuals or sampling units are nested within clusters, as is the case in EHR or IPD.  This hierarchical complexity must not only be taken into account in the analysis model, but also when dealing with missing data, thus requiring imputation models that are congenial to the analysis model, i.e. that make the same assumptions about the data.

Different procedures can be adopted to combine information between clusters; however, in our imputation method we opted for the two-stage approach that is often used in meta-analyses.[@simmonds2005] This is because such an approach is computationally less intensive and could potentially generate fewer convergence problems in the estimation of the Heckman hierarchical model compared to other approaches. Our method is based on the following conditional imputation model, whose parameters are $\boldsymbol{\theta}=\{\boldsymbol{\beta^O},\boldsymbol{\beta^S},\boldsymbol{\Psi^S},\boldsymbol{\Psi^O},(\sigma_i,\rho_i)_{1<=i<=N}\}$
\begin{align*}
\boldsymbol{y^*_{i}}&=\boldsymbol{X^O_{i}}(\boldsymbol{\beta^O}+b^O_{i})+\boldsymbol{\epsilon^O_{i}}\\
\boldsymbol{r^*_{i}}&=\boldsymbol{X^S_{i}}(\boldsymbol{\beta^S}+b^S_{i})+\boldsymbol{\epsilon^S_{i}}\\
b^O_{i}&\sim N(0,\boldsymbol{\Psi^O})\\
b^S_{i}&\sim N(0,\boldsymbol{\Psi^S})
\end{align*}
\[\begin{pmatrix}
\boldsymbol{\epsilon^O_{i}} \\
\boldsymbol{\epsilon^S_{i}}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho_i\sigma_i \\
\rho_i\sigma_i & 1
\end{pmatrix}\right)
\]

Briefly, in a first stage  we estimate the cluster specific parameters of the Heckman model $\boldsymbol{\theta_i}=\{\boldsymbol{\hat{\beta}_i^O},\boldsymbol{\hat{\beta}_i^S},\hat{\sigma}_i,\hat{\rho}_i\}$, for all $i$ between 1 and $N$ clusters.That is, we estimate separately for each of the clusters, the parameters of the two equations in each study, the outcome model (\ref{eq:1}) and the selection model (\ref{eq:2}) via a copula model.

In a second stage, we estimate random effects multivariate meta-analysis model for the coefficient parameters. Here $\boldsymbol{\beta_i^O}$ and $\boldsymbol{\beta_i^S}$parameters are assumed to be drawn independently and identically from a latent multivariate normal distribution of parameters with a  population means $\boldsymbol{\beta^O}$ and $\boldsymbol{\beta^S}$ and population variance covariance $\boldsymbol{\psi^O}$ and $\boldsymbol{\psi^S}$ respectively. [@higgins2009].Here we assume that $\sigma_i$ and $\rho_i$ are random variables that come from a populational distribution and we estimate random effects univariate meta-analysis models for these parameters.

# Using the Heckman model to impute missing data
\label{sec:methods}
We follow a similar approach proposed by Resche-Rigon and White (2018) [@resche-rigon2018] for multilevel data imputation, which allows us to impute values in very common scenarios in IPD, e.g., sporadic and systematic missingness patterns.  Our imputation method was created to impute datasets with a single missing outcome variable or covariate, but it can also be used to impute multiple incomplete variables in a dataset, as it can be implemented under a multivariate imputation by chain equations (MICE) approach.

@@

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\boldsymbol{\widehat{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. Here we obtain the estimates not only the parameters' point estimates $\boldsymbol{\widehat{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

@@

In this step, we pool the parameters $\boldsymbol{\widehat{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation. 
Thus, we can denote the study specific coefficients
$$\boldsymbol{\hat{\beta^O}_i} = \boldsymbol{\beta^O} +\boldsymbol{ b^O_i} + \boldsymbol{\epsilon'^O_{i}}$$
$$\boldsymbol{\hat{\beta^S}_i} = \boldsymbol{\beta^S} + b^S_i + \boldsymbol{\epsilon'^S_{i}}$$
using the random effects 
 $b^O_i \sim N(0,\boldsymbol{\Psi^O})$ and $b^S_i\sim N(0,\boldsymbol{\Psi^S})$ with $\epsilon'^O_{i}\sim N(0,\sigma^O_{i}(X^O_i{X^O_i}^T))$ and $\epsilon'^S_{i}\sim N(0,\sigma^S_{i}(X^S_i{X^S_i}^T))$. From the meta model we obtain the estimates of parameters $\boldsymbol{\beta^S}$,$\boldsymbol{\beta^O}$, $\boldsymbol{\Psi^S}$,$\boldsymbol{\Psi^O}$.

Assuming that  $\sigma_i$ and $\rho_i$ are random variables comming from a populational distribution, they can express as:
$$ log(\sigma_i) \sim N( log(\sigma),\psi^{\sigma})$$
$$ tanh^{-1}(\rho_i)\sim N( tanh^{-1}(\rho),\psi^{\rho})$$ 
For each of them, we perform a univariate random effects meta_analysis. The model for $\sigma_i$ is given by the model:
$$log(\hat\sigma_i) = \sigma + b^{\sigma}_i + \epsilon^{\sigma}_{i}$$
with $b_i^{\sigma} \sim N(0,\psi^{\sigma})$, and $\epsilon_i^{\sigma} \sim N(0,var(log(\hat{\sigma}_i)))$ and  the model for $\rho_i$ is:
$$tanh^{-1}(\hat\rho_i) = \rho + b^{\rho}_i + \epsilon^{\rho}_{i}$$
with $b_i^{\rho} \sim N(0,\psi^{\rho})$, and $\epsilon_i^{\rho} \sim N(0,var(tanh^{-1}(\hat{\rho}_i)))$.

@@

From the meta-analysis model, we obtain the marginal estimates $\boldsymbol{\widehat{\Theta}}=\{\widehat{\boldsymbol{\beta^S}},\widehat{\boldsymbol{\beta^O}},\hat{\sigma},\hat{\rho}\}$ and the between-cluster variance matrix $\boldsymbol{\widehat{\psi}}$,i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\Theta}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\Theta^*}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

@@

## Technical details of the implementation 
The Heckman model is estimated with the \textbf{gjrm} function of the GJRM R package[@radiceGJRMGeneralisedJoint2021],  under the bivariate model with the nonrandom sample selection (BSS) specification.
The meta-analysis model is performed with the \textbf{mixmeta} function of the R package mixmeta[@gasparrini2021], which allows the use of maximum likelihood (ML), restricted maximum likelihood (REML), and moments estimation methods. For the simulation and illustrative study, we use the restricted REML estimation method, which is recommended as it has a good balance between unbiasedness and efficiency. [@viechtbauer2005].

@@

- \textbf{Bias:} Bias on the coefficient and variance-covariance matrix terms.
- \textbf{Coverage:} Coverage of the 95\% confidence intervals for the coefficient  parameters.
- \textbf{Width}: Average width of the confidence interval on the coefficient parameters.
- \textbf{RMSE:} Root mean squared error of the coefficient  and random effect parameters.
